A Convolutional Neural Network for Outsole Recognition\tnoteref{t1}

### 2019-06-14 paper/paper.Rnw
A few changes/comments, updated images

Many convolutional neural networks have been trained on a \mt{sub}set of images \mt{from the database} ImageNet; these images are hierarchically labeled, so the same image of a dog may have labels of sky, grass, and trees, and the dog may be labeled as ``mammal'', ``dog'', ``retriever'', and ``golden retriever'' at increasing levels of label complexity \citep{dengImageNetLargeScaleHierarchical}. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. \fix{More to be said later.}

@@

\paragraph{VGG16}
Developed by Oxford's Visual Graphics Group, VGG16 is a \mt{pre-trained} CNN with 16 ``functional" (i.e., convolutional and densely connected) layers and 5 ``structural" max pooling layers \citep{krizhevskyImageNetClassificationDeep2012}. In contrast to other popular networks, like ResNet, VGG has a relatively simple structure that provides easier training and interpretability with very little sacrificed accuracy. The simplicity of this structure provides the ability to peer into the inner workings of the network for diagnostic purposes, providing a distinct advantage over more complicated network structures with slightly higher accuracy ratings. VGG16 is a common choice for transfer learning because of this structural simplicity; it has been used for detection of text in natural images \citep{zhongDeepTextUnifiedFramework2016}, medical imaging classification \citep{oquabLearningTransferringMidlevel2014}, classification of weld defects \citep{liuWeldDefectImages2018}, and many other domain-specific image recognition tasks that are more specific than the ImageNet data on which it was trained. \autoref{fig:VGG16-structure} shows the architecture of VGG16; with transfer learning, the VGG16 model head is replaced with a trained model head specific to recognition of shoeprint class characteristics.

\mt{The model base of VGG16 is made up of five convolutional blocks alternated with max pooling layers. Each convolutional block is made up of two or three convolutional layers, and each layer is made up of hundreds of filters, such that each of the five convolutional blocks has between 128 and 1,536 filters. Strictly speaking, a filter is a set of numerical weights that, when applied to an image, quantifies the presence of the specific shapes and/or colors that the filter is trained to detect. \fix{Necessary? (Or just skip...): Visualizing a filter then refers to visualizing the features to which the filter provides its maximum response.} \autoref{fig:vgg16layers} shows a selection of filters from VGG16, with each row corresponding to one of the five convolutional blocks. Filters in earlier convolutional blocks detect simple features, such as colors and blobs, while later filters detect more complex combinations of features. When considered together, the series of filters perform the process of feature detection, such that a single image is made up of some combination of the thousands of features that are detected by filters. Although the filters of VGG16 have been trained and optimized on the ImageNet Data, the large number of filters span a wide variety of shapes and features} \fix{which are still useful/relevant to classifying outsoles. In a sense, this is much like teaching humans to classify objects they have never seen before: an object may be unfamiliar, but our eyes can still detect its color, shape, and texture because we have learned these ideas from other sources.} \fix{Say somewhere: Max pooling layers aggregate the features in one block before passing them to the next block, which helps reduce the computational load of training the model and helps to generalize the relationship between features.}

@@

Receiver Operating Characteristic (ROC) curves are a type diagnostic plot which compare the false positive rate to the true positive rate for a classification method, and the Area Under the Curve (AUC) quantifies the shape of the ROC curve. Perfect prediction would be indicated by a right angle along the upper left corner of the plot, with a corresponding AUC of 1, and a diagonal line with an AUC of 0.5 would indicate that the classification method performs no better than random chance. \autoref{fig:overall-roc} shows the ROC curve for \fix{the model} across all classes, and \autoref{fig:class-roc} shows the curve for each class. The full model has an AUC of 0.88, and the AUC for individual classes ranges from 0.81 (for lines) to 0.91 (for bowtie and text). While the class performances do vary slightly, each ROC curve is the same general shape and performs significantly better than random chance. The points in \autoref{fig:class-roc} represent the equal error rate (EER), which is the threshold where there is equal probability of a false positive or false negative error. We will use the class-wise EER to set a threshold corresponding to each class, so that any model prediction with probability above the EER for that shape is considered a positive prediction.

@@

\paragraph{Model Predictions}
\mt{The ROC curves provide only a general summary of model accuracy; to learn more about how specific images are classified, we can view images from the test set along with their respective nine model predictions.} \autoref{fig:single-exemplars} shows model predictions for images that contain only one type of shape, and \autoref{fig:multi-exemplars} shows similar predictions for images containing multiple shapes. Most predictions for the single-label images are straightforward; all of the archetypical shapes are corectly identified by the model, with only small confusions like those between quad/polygon and circle/text. \fix{A later section will discuss possible causes for this confusion.}

@@

\fix{These figures provide representative examples of classifications across multiple categories; it is also useful to examine many cases within a given shape.} \mt{\autoref{fig:chevrons-consistency} shows a number of images containing chevrons. The top four rows show regularly repeating chevron patterns and seems to indicate that the model optimally identifies chevrons of a certain width; wide chevrons cross into the realm of bowties, whereas narrow chevrons become indistinguishable from lines. The next six rows show individual and closed chevrons. There is a significant amount of confusion between these wider chevrons and quadrilaterals. Finally, the bottom four rows show less-than-standard chevrons. The rounded chevron has a much lower prediction than those with sharper angles, but still exceeds the EER for chevron; the white chevrons are confounded with overlaid circles, which introduces a complexity that the model attributes to text.}

>>>>

Many convolutional neural networks have been trained on a subset of images \svp{from ImageNet}; these images are hierarchically labeled, so the same image of a dog may have labels of sky, grass, and trees, and the dog may be labeled as ``mammal'', ``dog'', ``retriever'', and ``golden retriever'' at increasing levels of label complexity \citep{dengImageNetLargeScaleHierarchical}. CNNs trained on ImageNet are optimized for general human-like vision, that is, the ability to recognize a large set of different features simultaneously. As a result, the base of networks trained on these sets are often used for transfer learning, because the initial layers are broadly generalizable to a wide variety of more specific image labeling tasks. \fix{More to be said later.}

@@

\paragraph{VGG16}
Developed by Oxford's Visual Graphics Group, VGG16 is a pre-trained CNN with 16 ``functional" (i.e., convolutional and densely connected) layers and 5 ``structural" max pooling layers \citep{krizhevskyImageNetClassificationDeep2012}. In contrast to other popular networks, like ResNet, VGG has a relatively simple structure that provides easier training and interpretability with very little sacrificed accuracy. The simplicity of this structure provides the ability to peer into the inner workings of the network for diagnostic purposes, providing a distinct advantage over more complicated network structures with slightly higher accuracy ratings. VGG16 is a common choice for transfer learning because of this structural simplicity; it has been used for detection of text in natural images \citep{zhongDeepTextUnifiedFramework2016}, medical imaging classification \citep{oquabLearningTransferringMidlevel2014}, classification of weld defects \citep{liuWeldDefectImages2018}, and many other domain-specific image recognition tasks that are more specific than the ImageNet data on which it was trained. \autoref{fig:VGG16-structure} shows the architecture of VGG16; with transfer learning, the VGG16 model head is replaced with a trained model head specific to recognition of shoeprint class characteristics.

\svp{The base of VGG16 is made up of five blocks, which consist of two or three convolutional layers and one max pooling layer which aggregates the signal from the convolutional layers. Each convolutional layer} is made up of hundreds of filters, such that each of the five convolutional blocks has between 128 and 1,536 filters. A filter is a set of numerical weights that, when applied to an image, quantifies the presence of the specific shapes and/or colors that the filter is trained to detect. \autoref{fig:vgg16layers} shows a selection of filters from VGG16, with each row corresponding to one of the five convolutional blocks. Filters in earlier convolutional blocks detect simple features, such as colors, \svp{lines, corners,} and blobs, while later filters detect more complex combinations of features. \svp{This aggregation of features through successive convolutional layers mirrors the process of complex feature detection in human vision.} \svp{Any complex feature can be disassembled into a set of simple features; similarly, the successive sets of layers detect increasingly detailed features.} %When considered together, the series of filters perform the process of feature detection, such that a single image is made up of some combination of the thousands of features that are detected by filters.
Although the filters of VGG16 have been trained and optimized on the ImageNet Data, the large number of filters span a wide variety of shapes and features, \svp{many of which are useful when examining tread patterns. Just as humans can detect the features of an unfamiliar object (consituent parts, edges, textures, and corners), a CNN trained on a wide set of objects can be used to classify unrelated images, such as shoe soles.} %\fix{which are still useful/relevant to classifying outsoles. In a sense, this is much like teaching humans to classify objects they have never seen before: an object may be unfamiliar, but our eyes can still detect its color, shape, and texture because we have learned these ideas from other sources.} \fix{Say somewhere: Max pooling layers aggregate the features in one block before passing them to the next block, which helps reduce the computational load of training the model and helps to generalize the relationship between features.} % I've tried to do this concisely - if you don't like it, please feel free to revert the changes.

@@

Receiver Operating Characteristic (ROC) curves are a type diagnostic plot which compare the false positive rate to the true positive rate for a classification method, and the Area Under the Curve (AUC) \svp{is a summary measure which} quantifies the \svp{overall accuracy of the model}. %It doesn't really summarize the shape - you can have very different shapes of curves with the same AUC value.
Perfect prediction would be indicated by a right angle along the upper left corner of the plot, with a corresponding AUC of 1, and a diagonal line with an AUC of 0.5 would indicate that the classification method performs no better than random chance. \autoref{fig:overall-roc} shows the ROC curve for \fix{the model} across all classes, and \autoref{fig:class-roc} shows the curve for each class. The full model has an AUC of 0.88, and the AUC for individual classes ranges from 0.81 (for lines) to 0.91 (for bowtie and text). \fix{pull these numbers out of the model using \\Sexpr} While the class performances do vary slightly, each ROC curve is the same general shape and performs significantly better than random chance. The points in \autoref{fig:class-roc} represent the equal error rate (EER), which is the threshold where there is equal probability of a false positive or false negative error. We will use the class-wise EER to set a threshold corresponding to each class, so that any model prediction with probability above the EER for that shape is considered a positive prediction.

@@

\paragraph{Model Predictions}
The ROC curves provide only a general summary of model accuracy; to learn more about how specific images are classified, we can examine images from the test set along with their respective nine model predictions. \autoref{fig:single-exemplars} shows model predictions for images that contain only one type of shape, and \autoref{fig:multi-exemplars} shows similar predictions for images containing multiple shapes. Most predictions for the single-label images are straightforward; all of the archetypical shapes are corectly identified by the model, with only small confusions like those between quad/polygon and circle/text. \fix{A later section will discuss possible causes for this confusion.}

@@

\fix{These figures provide representative examples of classifications across multiple categories; it is also useful to examine many cases within a given shape.} \autoref{fig:chevrons-consistency} shows a number of images containing chevrons. The top four rows show regularly repeating chevron patterns and seems to indicate that the model optimally identifies chevrons of a certain width; wide chevrons cross into the realm of bowties, whereas narrow chevrons become indistinguishable from lines. The next six rows show individual and closed chevrons. There is a significant amount of confusion between these wider chevrons and quadrilaterals. Finally, the bottom four rows show less-than-standard chevrons. The rounded chevron has a much lower prediction than those with sharper angles, but still exceeds the EER for chevron; the white chevrons are confounded with overlaid circles, which introduces a complexity that the model attributes to text.
